---
title: "Logistic Regression in Practice"
subtitle: Behavioral Data Science in R II
format: 
  revealjs:
    css: style.css
bibliography: ../references.bib
---

## Example: Extramarital affairs

We'll consider self-report data from 535 individuals with information about their family situation, views toward marriage, and whether or not they are currently having an extramarital sexual affair. The data comes from a survey conducted by Psychology Today in 1969.

## The data

```{r}
#| echo: true
library(tidyverse)
library(tidymodels)
library(broom)

affairs <- read_csv("data/affairs.csv") %>% 
  mutate(having_affair = factor(if_else(affairs > 0, 1, 0)))

head(affairs)
```

## Exploratory Data Analysis

```{r}
affairs %>% 
  group_by(rating) %>% 
  summarize(prop_affair = mean(having_affair==1), cnt = n()) %>% 
  mutate(se = sqrt(prop_affair*(1-prop_affair)/cnt)) %>% 
  ggplot(aes(x=rating, y = prop_affair, ymax = prop_affair + 2*se, ymin = prop_affair - 2*se )) +
  geom_pointrange() +
  ylim(0,1)
```


## Setting up

```{r}
#| echo: true
library(tidymodels)
set.seed(1234)
splits <- initial_split(affairs, strata = rating)

train <- training(splits)
test <- testing(splits)
```


## Simple logistic regression model

```{r}
lr_wflow <- workflow(having_affair ~ rating) %>% 
  add_model(logistic_reg())

lr_fit1 <- fit(lr_wflow, data = train)

lr_fit1
```


## Results

```{r}
tidy(lr_fit1)
```

## Plotting results

```{r}
lr_fit1 %>% 
  augment(
    tibble(rating = seq(1,5,.01))
  ) %>% 
  ggplot(aes(x = rating, y=.pred_1)) +
  geom_line() +
  geom_jitter(data = affairs, aes( x = rating, y = as.numeric(having_affair==1)), height = .05, alpha = .3)
```


## Adding predictors

## More visualizations

## Interpreting coefficients / predictions

# Performance metrics

## Accuracy

## Precision and Recall

## tidymodels vs `glm()`




