---
title: "Interpretability and Feature Importance"
subtitle: Behavioral Data Science in R II
format: 
  revealjs:
    css: style.css
bibliography: ../references.bib
---

## Interpretability tools

We need ways to look inside the black box

- Feature importance
- Diagnostic plots
  - Partial Dependency Plots
  - Individual Conditional Expectation Plots



```{r}
#| message: false
#| include: false
library(tidyverse)
library(tidymodels)
library(hutils)
library(DALEX)
library(DALEXtra)
taxi <- read_csv("data/nyc-taxi.csv") %>% 
  mutate(pickup_datetime = lubridate::as_datetime(pickup_datetime))
```

```{r}
#| include: false
splits <- initial_split(taxi, prop = .8)
train <- training(splits)
test <- testing(splits)

folds <- vfold_cv(train, 10)
```

```{r}
#| include: false
train %>% 
  ggplot(aes(x=pickup_longitude, y = pickup_latitude)) +
  geom_point(alpha = .2, size = .2, shape = 16, color = "goldenrod1") +
  theme_void() +
  theme(panel.background = element_rect(fill = "black"))
```


```{r}
#| include: false
fe_rec <- recipe(trip_duration ~ ., train) %>%
  step_mutate(
    dist = haversine_distance(pickup_latitude, pickup_longitude, dropoff_latitude, dropoff_longitude),
    bearing = bearing(pickup_latitude, pickup_longitude, dropoff_latitude, dropoff_longitude)
    ) %>%
  step_date(pickup_datetime, features = c("dow", "doy")) %>%
  step_time(pickup_datetime, features = c("decimal_day")) %>% 
  update_role(pickup_datetime, new_role = "id")

train_juiced <- juice(prep(fe_rec))
```

```{r}
#| include: false
rf_mod <- rand_forest(mode = "regression")

wf <- workflow() %>% 
  add_recipe(fe_rec) %>% 
  add_model(rf_mod)

fitted_model <- wf %>% 
  fit(train)

```


```{r}
#| include: false
make_explainer_obj <- function(fitted_workflow){
  fitted_model <-
    fitted_workflow %>% 
    extract_fit_parsnip() # <- parsnip model_fit object

  feature_data <- 
    fitted_workflow %>% 
    extract_mold() %>% 
    pluck("predictors") 
  
  outcome_data <- 
    fitted_workflow %>% 
    extract_mold() %>% 
    pluck("outcomes") %>% 
    pluck(1)    # <- is is a 1D df, make it a vector
  
  vip_features <- 
    explain_tidymodels(
      fitted_model, 
      data = feature_data, 
      y = outcome_data
    )
  
  return(vip_features)
}

```

## Setting up an explainer

```{r}
#| message: false
library(DALEX)
library(DALEXtra)

explainer_rf <- make_explainer_obj(fitted_model)
```


## PDP plots

Partial dependence plots visualize the **marginal** influence of a predictor, averaging over the other predictors.

::: {.callout-tip icon="false"}
## Algorithm
For a selected predictor (x)
1. Construct a grid of j evenly spaced values across the distribution
   of $x$: ${x1, x2, ..., xj}$
2. For $i$ in ${1,...,j}$ do
     | Copy the training data and replace the original values of x 
       with the constant xi
     | Apply given ML model (i.e., obtain vector of predictions)
     | Average predictions together
   End
3. Plot the averaged predictions against x1, x2, ..., xj
:::

## Visualizing time-of-day effects

```{r}
#| echo: false

# train_juiced %>% 
#   ggplot(aes(x = pickup_datetime_dow, y = trip_duration)) +
#     geom_boxplot(size = .1)

# train_juiced %>% 
#   mutate(speed_kmph = dist / exp(trip_duration) * 3600) %>% 
#   mutate(pickup_datetime_decimal_day = round(pickup_datetime_decimal_day, 1)) %>% 
#   group_by(pickup_datetime_dow) %>% 
#   summarize(M = mean(trip_duration), se = sd(trip_duration)/sqrt(n()))  %>% 
#   ggplot(aes(x = pickup_datetime_dow, y = M)) +
#   geom_pointrange(aes(ymin = M - 2 * se, ymax = M + 2*se))

# train_juiced %>% 
#   mutate(speed_kmph = dist / exp(trip_duration) * 3600) %>% 
#   mutate(pickup_datetime_decimal_day = round(pickup_datetime_decimal_day, 1)) %>% 
#   group_by(pickup_datetime_decimal_day) %>% 
#   summarize(M = mean(trip_duration), se = sd(trip_duration)/sqrt(n()))  %>% 
#   ggplot(aes(x = pickup_datetime_decimal_day, y = M)) +
#   geom_line() +
#   geom_ribbon(aes(ymin = M - 2 * se, ymax = M + 2*se), alpha = .2)
```


```{r}
pdp_time <- model_profile(explainer_rf, N = 100, variables = "pickup_datetime_decimal_day")
plot(pdp_time)

```

## Individual Conditionalized Expectation

- ICE plots visualize the effect of a predictor for a sample of individual cases.
- Help to visualize heterogeneity in predictor effects (i.e., does predictor effect vary depending on levels of other variables)

## ICE plots

```{r}

ggplot_ice <- function(obj, x) {
  agg_d <- as_tibble(obj$agr_profiles) %>%
      mutate(`_label_` = stringr::str_remove(`_label_`, "^[^_]*_")) 
  
  first_y <- agg_d %>% summarize(first_y = first(`_yhat_`)) %>% pull(first_y)
  
  p <- 
    agg_d %>% 
    mutate(`_yhat_` = `_yhat_` - first_y) %>% 
    ggplot(aes(`_x_`, `_yhat_`)) +
    geom_line(data = as_tibble(obj$cp_profiles) %>% 
                group_by(`_ids_`) %>% 
                mutate(firsts = first(`_yhat_`)) %>% 
                mutate(`_yhat_` = `_yhat_` - firsts),
              aes(x = {{ x }}, group = `_ids_`),
              linewidth = 0.5, alpha = 0.1, color = "gray50")
  
  num_colors <- n_distinct(obj$agr_profiles$`_label_`)
  
  if (num_colors > 1) {
    p <- p + geom_line(aes(color = `_label_`), linewidth = 1.2, alpha = 0.8)
  } else {
    p <- p + geom_line(color = "midnightblue", linewidth = 1.2, alpha = 0.8)
  }
  
  p
}

ggplot_ice(pdp_time, pickup_datetime_decimal_day)
```


## Visualizing distance and `trip_duration`

```{r}
pdp_dist <- model_profile(explainer_rf, N = 100, variables = "dist")
plot(pdp_dist)
```

```{r}
ggplot_ice(pdp_dist, dist)
```

```{r}
pdp_dist <- model_profile(explainer_rf, N = 100, variables = "bearing")
ggplot_ice(pdp_dist, bearing)
```

```{r}
pdp_2d <- model_profile(explainer_rf, N = 100, variables = c("bearing", "pickup_datetime_decimal_day"), type = "conditional")
```

```{r}
Nsteps <- 100

grid <- crossing(
  bearing = seq(0, 360, length.out = Nsteps),
  pickup_datetime_decimal_day = seq(min(juiced_df$pickup_datetime_decimal_day), max(juiced_df$pickup_datetime_decimal_day), length.out = Nsteps),
)

pred_df <-  fitted_model %>% 
    extract_mold() %>% 
    pluck("predictors")  %>% 
  sample_n(11) %>% 
  mutate(.id = 1:n()) %>% 
  select(-bearing, -pickup_datetime_decimal_day) %>% 
  cross_join(grid)

extract_fit_parsnip(fitted_model) %>% 
  augment(pred_df) %>% 
  group_by(bearing, pickup_datetime_decimal_day) %>% 
  summarize(m = mean(.pred)) %>% 
  ggplot(aes(x=bearing, y = pickup_datetime_decimal_day, fill = m)) +
  geom_tile(color = NA) +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(panel.grid = element_blank())

```

```{r}
# library(plotly)
# pred_2d <- extract_fit_parsnip(fitted_model) %>%
#   augment(pred_df) %>%
#   group_by(dist, pickup_datetime_decimal_day) %>%
#   summarize(m = mean(.pred)) %>%
#   ungroup() %>%
#   as.matrix()
# 
# fig <- plot_ly(z = ~pred_2d)
# 
# fig <- fig %>% add_surface()
# 
# 
# fig

```

